---
title: "Northeastern States Correlations"
output:
  html_document:
    df_print: paged
---
In this markdown I will look at each of the features and run correlations on them to see how they relate with poverty proportion. I will also graph out each one of them to get a visual on the relationship.



I am going to first load in the project data that was prepared and then take a look at it.

```{r}
library(tidyverse)
project_data <- read_csv("northeast_data.csv")
project_data
```

Look at a plot of poverty compared to new teacher.

```{r}
project_data %>%
  ggplot(aes(New_Teachers_Proportion, Poverty_Proportion)) + 
  geom_point()
```
I am going to look at the correlation for new teachers.

```{r}
cor.test(project_data$New_Teachers_Proportion, project_data$Poverty_Proportion)
```
I am going to graph out absent teacher ratio to see the relationship.

```{r}
project_data %>%
  ggplot(aes(Absent_Teacher_Proportion, Poverty_Proportion)) + 
  geom_point()
```
Looking at the correlation between for absent teachers.

```{r}
cor.test(project_data$Absent_Teacher_Proportion, project_data$Poverty_Proportion)
```
Looking at the graph of counselor ratio.

```{r}
project_data %>%
  ggplot(aes(Counselor_Ratio, Poverty_Proportion)) + 
  geom_point()
```
Look at the correlation for counselor ratio.

```{r}
cor.test(project_data$Counselor_Ratio, project_data$Poverty_Proportion)
```
Looking at the graph with the student_teacher_ratio.

```{r}
project_data %>%
  ggplot(aes(Student_Teacher_Ratio, Poverty_Proportion)) + 
  geom_point()
```
Looking at the correlation with student teacher ratio.

```{r}
cor.test(project_data$Student_Teacher_Ratio, project_data$Poverty_Proportion)
```
Looking at how chronic abseentism graphs with poverty.

```{r}
project_data %>%
  ggplot(aes(Absent_Prop, Poverty_Proportion)) + 
  geom_point()
```
Looking at the correlation for absent and poverty.

```{r}
cor.test(project_data$Absent_Prop, project_data$Poverty_Proportion)
```
Looking at the graph of students that have taken the SAT or ACT compared to poverty.

```{r}
project_data %>%
  ggplot(aes(Test_Prop, Poverty_Proportion)) + 
  geom_point()
```
Look at the correlation of test takers.

```{r}
cor.test(project_data$Test_Prop, project_data$Poverty_Proportion)
```
Looking at how sports team relates to poverty.

```{r}
project_data %>%
  ggplot(aes(Teams_Prop, Poverty_Proportion)) + 
  geom_point()
```
Looking at the correlation of teams to poverty.

```{r}
cor.test(project_data$Teams_Prop, project_data$Poverty_Proportion)
```
Looking at athletes compared to poverty.

```{r}
project_data %>%
  ggplot(aes(Athletes_Prop, Poverty_Proportion)) + 
  geom_point()
```
Looking at the correlation of athletes.

```{r}
cor.test(project_data$Athletes_Prop, project_data$Poverty_Proportion)
```
Looking at how ap test takers compare to poverty.

```{r}
project_data %>%
  ggplot(aes(Takers_Proportion, Poverty_Proportion)) + 
  geom_point()
```
Looking at the correlation for the Ap test takers.

```{r}
cor.test(project_data$Takers_Proportion, project_data$Poverty_Proportion)
```
Looking at AP students compared to poverty.

```{r}
project_data %>%
  ggplot(aes(AP_Prop, Poverty_Proportion)) + 
  geom_point()
```
Correlation of AP students to poverty.

```{r}
cor.test(project_data$AP_Prop, project_data$Poverty_Proportion)
```
Suspension days compared to poverty.

```{r}
project_data %>%
  ggplot(aes(AVG_Suspenion, Poverty_Proportion)) + 
  geom_point()
```
I want to look at the correlation of suspension with poverty.

```{r}
cor.test(project_data$AVG_Suspenion, project_data$Poverty_Proportion)
```
I want to look at expulsions compared to poverty.

```{r}
project_data %>%
  ggplot(aes(Expulsion_Prop, Poverty_Proportion)) + 
  geom_point()
```
Look at correlation of expulsion to poverty.

```{r}
cor.test(project_data$Expulsion_Prop, project_data$Poverty_Proportion)
```
Police involvement compared to poverty.

```{r}
project_data %>%
  ggplot(aes(Police_Prop, Poverty_Proportion)) + 
  geom_point()
```
Correlation of police involvement to poverty.

```{r}
cor.test(project_data$Police_Prop, project_data$Poverty_Proportion)
```
Student proportion compared to population and poverty.

```{r}
project_data %>%
  ggplot(aes(Student_Prop, Poverty_Proportion)) + 
  geom_point()
```
Look at the correlation of the student proportion.

```{r}
cor.test(project_data$Student_Prop, project_data$Poverty_Proportion)
```
For the next part I will look at correlations as they relate to bullying on race and gender.

```{r}
project_data %>%
  ggplot(aes(Race_Disc, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Race_Disc, project_data$Poverty_Proportion)
```
```{r}
project_data %>%
  ggplot(aes(Race_Reports, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Race_Reports, project_data$Poverty_Proportion)
```
```{r}
project_data %>%
  ggplot(aes(Gender_Disc, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Gender_Disc, project_data$Poverty_Proportion)
```
```{r}
project_data %>%
  ggplot(aes(Gender_Reports, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Gender_Reports, project_data$Poverty_Proportion)
```
I will now run correlations on the different courses that are offered by districts.

```{r}
project_data %>%
  ggplot(aes(Alg1, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Alg1, project_data$Poverty_Proportion)
```
```{r}
project_data %>%
  ggplot(aes(Alg2, Poverty_Proportion)) + 
  geom_point()
```

```{r}
cor.test(project_data$Alg2, project_data$Poverty_Proportion)
```
```{r}
project_data %>%
  ggplot(aes(Geo, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Geo, project_data$Poverty_Proportion)
```
```{r}
project_data %>%
  ggplot(aes(AdvMath, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$AdvMath, project_data$Poverty_Proportion)
```
```{r}
project_data %>%
  ggplot(aes(Calc, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Calc, project_data$Poverty_Proportion)
```
```{r}
project_data %>%
  ggplot(aes(Bio, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Bio, project_data$Poverty_Proportion)
```
```{r}
project_data %>%
  ggplot(aes(Chem, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Chem, project_data$Poverty_Proportion)
```
```{r}
project_data %>%
  ggplot(aes(Phys, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Phys, project_data$Poverty_Proportion)
```
I am going to look at the data for students that passed Algebra and the correlations.

```{r}
project_data %>%
  ggplot(aes(Early_Pass, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Early_Pass, project_data$Poverty_Proportion)
```
```{r}
project_data %>%
  ggplot(aes(Late_Pass, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Late_Pass, project_data$Poverty_Proportion)
```
Look at demographics and how they relate to poverty.

```{r}
project_data %>%
  ggplot(aes(Non_White_Students, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Non_White_Students, project_data$Poverty_Proportion)
```



Look at how grad rate and poverty relate and the correlation of it.

```{r}
project_data %>%
  ggplot(aes(Grad_Rate, Poverty_Proportion)) + 
  geom_point()
```
```{r}
cor.test(project_data$Grad_Rate, project_data$Poverty_Proportion)
```

